Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
fix breakage

For many concurrent data structures, the ideal correctness condition requires that the data structure behaves identically to a sequential implementation, even when accessed concurrently by multiple threads. The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}. These take the form of $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple suggests that if $\texttt{P}_l$ holds true before a call and $\texttt{P}_p$ is true for some value of $a$ in a shared state (e.g. a global invariant), then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends). For instance, the specification
$\forall s.\ \left\langle \texttt{is\_stack}\ p\ |\ \texttt{stack}\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}\ p\ |\ \texttt{stack}\ (v::s)\right\rangle$
expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution. The stack itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the \texttt{is\_stack} assertion.

To demonstrate the proof of an implementation that meets the atomic specification, let's take the example of the stack specification mentioned above. This specification assert that the implementation \texttt{push} appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states. The local preconditions and postconditions, \texttt{is\_stack}, contain the points-to and lock-invariant assertions necessary for a thread, while the public preconditions and postconditions, \texttt{stack}, involve the ghost state that defines the overall state of the stack. Ghost state is a type of program state that aids in the verification process without affecting the program's runtime behavior, and by referring to an arbitrary ghost state identifier, it is possible to connect local and public conditions. To show that the implementation of \texttt{push}(v) satisfies these atomic specifications, two things must be demonstrated. Firstly, the implementation can execute safely without owning any piece of public assertion $\texttt{stack}$, only accessing it automatically and restoring it up until the linearization point. Secondly, at some linearization point, the implementation transforms the current stack $s$ into $v::s$ atomically, which satisfies the public postcondition $\texttt{stack}(v::s)$.

@@

\begin{mathpar}
	% spec of findNext
	{\color{blue}\{ \  \texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} \ \}}
	\\ 
	\texttt{findNext(pn, k)} 
	\\
	{\color{blue} \{ \texttt{res n'.} \ \texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)}  \ast }
	\\ 
	{\color{blue} \texttt{if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} 
	\texttt{ else } \texttt{(n' = n} \land \texttt{k = k')}  \}}
	\\
	\\
	%spec of insertOp
	{\color{blue} \left\{ \texttt{pn->n} \ast \texttt{n->t}  \right\}  } 
	\\
	\texttt{insertOp(pn, k, v)}
	\\
	{\color{blue} \{\texttt{t' l r lk1 lk2}. \ \ \texttt{pn->n} \ast \texttt{n->t'} \ast \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ast }
	\\
	{\color{blue} \texttt{l} \mapsto \texttt{(NULL, lk1)} \ast \texttt{r} \mapsto \texttt{(NULL, lk2)}
	\} }
\end{mathpar}

The specification of \texttt{findNext} states that a non-null \texttt{t} field points to a \texttt{node} struct containing the current node's key  \texttt{k'}, the node's value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes. When \texttt{findNext} sucessfully locates the next node \texttt{n'}, it sets \texttt{n'} as the left child node \texttt{l} if the provided key \texttt{k} is less than the current node's key \texttt{k'}. Conversely, if the given key \texttt{k} is greater than the current node's key \texttt{k'}, the next node \texttt{n'} is set as the right child node \texttt{r}. In case where \texttt{findNext} is unable to identify the next node \texttt{n'}, it retains the current node \texttt{n} as \texttt{n'},  indicating no change and that the provided key \texttt{k} is  identical to the current node's key \texttt{k'}.

The \texttt{insertOp} specification indicates that once \texttt{traverse} successfully reaches a leaf node (either the left or right node), the leaf node will have its \texttt{t} field set to \texttt{NULL}. This signifies that the leaf node currently has no key or value. At this location, a new node with the specified key \texttt{k} and value \texttt{v} should be inserted. Upon locating the appropriate leaf node for adding the new key-value pair \texttt{(k, v)}, the process generates two new empty child nodes with corresponding locks, \texttt{(NULL, lk1)} and \texttt{(NULL, lk2)}, which are pointed to by the \texttt{l} and \texttt{r} of the current node.

@@

>>>>

For many concurrent data structures, the ideal correctness condition requires that the data structure behaves identically to a sequential implementation, even when accessed concurrently by multiple threads. The concurrent behavior of search structures is specified using \emph{atomic triples}~\cite{tada}. These take the form of $\forall a.\ \left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple suggests that if $\texttt{P}_l$ holds true before a call and $\texttt{P}_p$ is true for some value of $a$ in a shared state (e.g. a global invariant), then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends). For instance, the specification
$\forall s.\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ s\right\rangle\ \texttt{push}(v)\ \left\langle \texttt{is\_stack}_g\ p\ |\ \texttt{stack}_g\ (v::s)\right\rangle$
expresses the fact that the push operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from particular stack $s$ to $v::s$ at some point during its execution. The stack itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the \texttt{is\_stack} assertion.

To demonstrate the proof of an implementation that meets the atomic specification, let's take the example of the stack specification mentioned above. This specification assert that the implementation \texttt{push} appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states. The local preconditions and postconditions, \texttt{is\_stack}, contain the points-to and lock-invariant assertions necessary for a thread, while the public preconditions and postconditions, \texttt{stack}, involve the ghost state that defines the overall state of the stack. Ghost state is a type of program state that aids in the verification process without affecting the program's runtime behavior, and by referring to an arbitrary ghost state identifier $g$ (which we will generally omit), it is possible to connect local and public conditions. To show that the implementation of \texttt{push}(v) satisfies these atomic specifications, two things must be demonstrated. Firstly, the implementation can execute safely without owning any piece of public assertion $\texttt{stack}$, only accessing it automatically and restoring it up until the linearization point. Secondly, at some linearization point, the implementation transforms the current stack $s$ into $v::s$ atomically, which satisfies the public postcondition $\texttt{stack}(v::s)$.

@@

\begin{figure}
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{findNext(pn, k)} 
		\\
		{\color{blue}
			\left\{\exists \  \texttt{res n'.}
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{t} \mapsto \texttt{(k', v', l, r)} \  \ast \\ 
				\ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n' = n} \land \texttt{k = k'))}
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the lock-coupling and give-up templates}
	\label{fig:findNext_lock}
\end{figure}

\begin{figure}
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\{\exists \  \texttt{t' l r lk1 lk2}.
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
				\ \texttt{l} \mapsto \texttt{(NULL, lk1)}\  \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2)}
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{insertOp} for the lock-coupling template}
	\label{fig:insertOp_lock}
\end{figure}

The specification of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that a non-null \texttt{t} field points to a \texttt{node} struct containing the current node's key  \texttt{k'}, the node's value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes. When \texttt{findNext} sucessfully locates the next node \texttt{n'}, it sets \texttt{n'} as the left child node \texttt{l} if the provided key \texttt{k} is less than the current node's key \texttt{k'}. Conversely, if the given key \texttt{k} is greater than the current node's key \texttt{k'}, the next node \texttt{n'} is set as the right child node \texttt{r}. In case where \texttt{findNext} is unable to identify the next node \texttt{n'}, it retains the current node \texttt{n} as \texttt{n'},  indicating no change and that the provided key \texttt{k} is  identical to the current node's key \texttt{k'}.

The \texttt{insertOp} specification (Figure \ref{fig:insertOp_lock}) indicates that once \texttt{traverse} successfully reaches a leaf node (either the left or right node), the leaf node will have its \texttt{t} field set to \texttt{NULL}. This signifies that the leaf node currently has no key or value. At this location, a new node with the specified key \texttt{k} and value \texttt{v} should be inserted. Upon locating the appropriate leaf node for adding the new key-value pair \texttt{(k, v)}, the process generates two new empty child nodes with corresponding locks, \texttt{(NULL, lk1)} and \texttt{(NULL, lk2)}, which are pointed to by the \texttt{l} and \texttt{r} of the current node.

@@

\begin{figure}
	\centering
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n} \ \ast \texttt{n->t} \ \ast \ \texttt{n->min} \ \ast \ \texttt{n->max}
		\end{array}
		\right\}
	}
	\\ 
	\texttt{inRange(pn, k)} 
	\\
	{\color{blue}
		\left\{\exists \  \texttt{res.}
		\begin{array}{c}
			 \ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{n->min} \ \ast \ \texttt{n->max}\  \ast \\ 
			\ \texttt{(if res then } \texttt{(min < x} \land \texttt{x < max)} \land \texttt{(x} \leq \texttt{min} \lor \texttt{x} \geq \texttt{max))}
		\end{array}
		\right\}
	}
\end{mathpar}
\caption{Specification of \texttt{inRange} for the give-up template}
\label{fig:inRange_giveup}
\end{figure}

\begin{figure}
	\centering
	\begin{mathpar}
	%spec of insertOp
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n} \ \ast \texttt{n->t} \  \ast \texttt{n->min} \ \ast \texttt{n->max}
		\end{array}
		\right\}
	}
	\\
	\texttt{insertOp(pn, k, v)}
	\\
	{\color{blue}
	\left\{\exists \ \texttt{t' l r lk1 lk2}. 
	\begin{array}{l}
		\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\
		\ \texttt{l} \mapsto \texttt{(NULL, lk1, (min, x))} \ \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2, (x, max))}
	\end{array}
	\right\}
	}
\end{mathpar}
\caption{Specification of \texttt{insertOp} for the give-up template}
\label{fig:insertOp_giveup}
\end{figure}
